### Ecological niche modeling using kuenm

The following code was created by (Cobos et al., 2019).

For more details about installing the kuenm package and about the organization of the data it is recommended to see the publication by (Cobos et al., 2019) and the github repository https://github.com/marlonecobos/kuenm

Reference
Cobos ME, Peterson AT, Barve N, Osorio-Olvera L. 2019. kuenm: an R package for detailed development of ecological niche models using Maxent. Peer J 7: e6281. https://doi.org/10.7717/peerj.6281


```{r, eval=FALSE}
setwd()# Set working directory

library(kuenm)
```

<br>

### Pre-modeling 

## Training and testing data splitting. Randomly 75% for training and 25% for testing.

```{r, eval=FALSE}
occ_thinn <- read.csv("Occurrences after spatial thinning.csv")
occ_thinn$name <- gsub(" ", "_", occ_thinn$name)
all <- unique(occ_thinn)

all$check <- paste(all[,2], all[,3], sep = "_")
train <- all[sample(nrow(all), round((length(all[,1])/4 *3))), ]
test <- all[!all[,4] %in% train[,4], ]

all$check <- NULL
train$check <- NULL
test$check <- NULL

write.csv(all, "spname_joint.csv", row.names = FALSE)
write.csv(train, "spname_train.csv", row.names = FALSE)
write.csv(test, "spname_test.csv", row.names = FALSE)
```

<br>

## Evaluation and plot of variable contribution to Maxent models

```{r, eval=FALSE}
help(explore_var_contrib)
```

<br>

```{r, eval=FALSE}

occ <- read.csv("spname_joint.csv")

mvars <- raster::stack(list.files(path ="Directory where Set1 is located.../M_variables/Set1", 
                         pattern = ".asc$", full.names = T))# For Windows users it is necessary to change the symbols "\" to "/"
                                
var_cont <- explore_var_contrib(occ = occ, M_variables = mvars, maxent.path = "Directory where the MaxEnt program is located", plot = FALSE)

plot_contribution(var_cont)
```

<br>

### Model calibration

## Candidate models

Candidate models are a large set of candidate models created to respond to the need to test broad suites of parameter combinations, such as, distinct regularization multiplier values, various feature classes, and different sets of environmental variables. The following code calls the help page of the function kuenm_cal.

```{r, eval=FALSE}
help(kuenm_cal)
```

<br>

The next chunk of code is for preparing the arguments to use the function. These variables should be changed according to each case.

```{r, eval=FALSE}
occ_joint <- "spname_joint.csv"
occ_tra <- "spname_train.csv"
M_var_dir <- "M_variables"
batch_cal <- "Candidate_models"
out_dir <- "Candidate_Models"
reg_mult <- c(seq(0.1, 1, 0.1), seq(2, 6, 1), 8, 10)
f_clas <- "all"
MB <- 100000
threads <- 3
args <- NULL
maxent_path <- "Directory where the MaxEnt program is located"
wait <- FALSE
run <- TRUE
```

<br>

The following is the code for using the function.

```{r, eval=FALSE}
kuenm_cal(occ.joint = occ_joint, occ.tra = occ_tra, M.var.dir = M_var_dir, 
          batch = batch_cal, out.dir = out_dir, reg.mult = reg_mult, 
          f.clas = f_clas, args = args, maxent.path = maxent_path, 
          wait = wait, run = run)
```

<br>

## Evaluation and selection of best models

Evaluation is a crucial step in model calibration. This step centers on selecting candidate models and their associated parameters to identify the best models for the purposes of the study. The kuenm_ceval function evaluates candidate models based on three distinct criteria: statistical significance (based on partial ROC analyses), prediction ability (omission rates), and model fit and complexity (using AICc). The following code chunk calls the function help window.

```{r, eval=FALSE}
help(kuenm_ceval)
```

<br>

Below, arguments for this functions will be defined. Change them as needed.

```{r, eval=FALSE}
occ_test <- "spname_test.csv"
out_eval <- "Calibration_results"
threshold <- 5
rand_percent <- 50
iterations <- 500
kept <- T
selection <- "OR_AICc"
parallel.proc <- T
# Note, some of the variables used here as arguments were already created for the previous function
```

<br>

This code allows evaluating candidate models that were created previously, selecting those with best performance based on the three criteria.

```{r, eval=FALSE}
cal_eval <- kuenm_ceval(path = out_dir, occ.joint = occ_joint, occ.tra = occ_tra, 
                        occ.test = occ_test, batch = batch_cal, out.eval = out_eval, 
                        threshold = threshold, rand.percent = rand_percent, 
                        iterations = iterations, kept = kept, selection = selection)
```

<br>

### Final model(s) 

## Final model creation

After selecting parameterizations that produce best models, the next step is to create the final models, and if needed transfer them to other environmental data sets (e.g., to other time periods or other geographic regions). The help page of kuenm_mod is called via this code:

```{r, eval=FALSE}
help(kuenm_mod)
```

<br>

To prepare the arguments for this function use the following chunk of code. Change variables as needed.

```{r, eval=FALSE}
batch_fin <- "Final_models"
mod_dir <- "Final_Models"
rep_n <- 10
rep_type <- "Bootstrap"
max.memory <- 10000
jackknife <- FALSE
out_format <- "cloglog"
project <- TRUE
G_var_dir <- "G_variables"
ext_type <- "ext"
MB <- 100000
threads <- 3
write_mess <- FALSE
write_clamp <- FALSE
wait1 <- FALSE
run1 <- TRUE
args <- NULL
# Again, some of the variables used as arguments were already created for the previous functions
```

<br>

The kuenm_mod function has the following syntax:

```{r, eval=FALSE}
kuenm_mod(occ.joint = occ_joint, M.var.dir = M_var_dir, out.eval = out_eval, 
          batch = batch_fin, rep.n = rep_n, rep.type = rep_type, 
          jackknife = jackknife, out.dir = mod_dir, out.format = out_format, 
          project = project, G.var.dir = G_var_dir, ext.type = ext_type,
          write.mess = write_mess, write.clamp = write_clamp, 
          maxent.path = maxent_path, args = args, wait = wait1, run = run1)
```

<br>

### Evaluation of extrapolation risks

If transfers were performed when creating final models, the MOP analysis will help to identify areas of strict extrapolation and levels of similarity between the calibration area and the region or scenario of projection. The code below will help to see the function's documentation:

```{r, eval=FALSE}
help(kuenm_mmop)
```

<br>

Below, arguments for this functions will be defined. Change them as needed.

```{r, eval=FALSE}
is_swd <- FALSE
sets_var <- c("Set2")
out_mop <- "MOP_results"
percent <- 10
paral <- T # make this true to perform MOP calculations in parallel, recommended
               # only if a powerful computer is used (see function's help)
# Some of the variables used as arguments were already created for the previous functions
```

<br>

The kuenm_mmop function has the following syntax:

```{r, eval=FALSE}
kuenm_mmop(G.var.dir = G_var_dir, M.var.dir = M_var_dir, is.swd = is_swd,
           sets.var = sets_var, out.mop = out_mop, percent = percent, 
           parallel = paral)
```

<br>

### Post-modeling analysis

Other functions in kuenm allow to perform analyses that summarize results and help with interpretations. These functions are `kuenm_modstats`, `kuenm_projchanges`, `kuenm_modvar`, `kuenm_hierpart`, and `kuenm_mopagree`. More details on how to use these functions can be found <a href="https://github.com/marlonecobos/kuenm/blob/master/extra_vignettes/post-modeling.md#post-modeling-analyses-with-kuenm" target="_blank">here</a>  

#Calculation of descriptive statistics of models

```{r, eval=FALSE}
help(kuenm_modstats)
```

<br>

##Preparing arguments:

```{r, eval=FALSE}
sp_name <- "Species_name" #Use the species name as written in the first column of your "spname_joint.cvs" file.
fmod_dir <- "Final_Models"
format <- "asc"
project <- TRUE
stats <- c("med", "range")
rep <- TRUE
scenarios <- c("aacurr", "acc_126", "acc_585", "aus", "can_126", "can_585", "gis_126", "gis_585")
ext_type <- c("E") # the type of extrapolation can be selected according to user requirements
out_dir <- "Final_Model_Stats" 
```

<br>

# argument "time.periods" is not included in the example but it can be used when models
# are projected to more than one time period, other than current.

##The following lines will execute the function:

```{r, eval=FALSE}
kuenm_modstats(sp.name = sp_name, fmod.dir = fmod_dir, format = format, project = project, 
               statistics = stats, replicated = rep, proj.scenarios = scenarios, 
               ext.type = ext_type, out.dir = out_dir)
```

<br>

#Detection of changes in models projected in time         

```{r, eval=FALSE}
help(kuenm_projchanges)
```

<br>

### other arguments were defined before

```{r, eval=FALSE}
occ <- "spname_joint.csv"
fmod_stats <- "Final_Model_Stats"
thres <- 5
curr <- "aacurr"
emi_scenarios <- c("126", "585")
c_mods <- c("acc", "can", "gis")
ext_type <- c("E")
out_dir1 <- "Projection_Changes"
```

<br>

##The following lines will execute the function:

```{r, eval=FALSE}
kuenm_projchanges(occ = occ, fmod.stats = fmod_stats, threshold = thres, current = curr, 
                  emi.scenarios = emi_scenarios, clim.models = c_mods, ext.type = ext_type, 
                  out.dir = out_dir1)
```

<br>
          
#Mop agree: if multiple GCMs where used, users may find convenient to use the kuenm_mopagree function, which creates raster layers of strict extrapolative areas agreement among distinct GCMs.

```{r, eval=FALSE}
help(kuenm_mopagree)
```

<br>

# MOP layers must be already created before using this function.

# Arguments

```{r, eval=FALSE}
mop_dir <- "MOP_results"
format <- "GTiff"
curr <- "aacurr"
emi_scenarios <- c("126", "585")
out_dir <- "MOP_agremment"
```

<br>

#Run MOP agremment

```{r, eval=FALSE}
kuenm_mopagree(mop.dir = mop_dir, in.format = format, out.format = format,
               current = curr,
               emi.scenarios = emi_scenarios, out.dir = out_dir)
```

<br>


